Highly Available Transactions: Virtues and Limitations

### 2013-04-23 writeup/intro.tex
editing pass

The prior decade has seen a large shift in the design of large-scale
database backends, from the use of transactional
RDBMSs~\cite{bernstein-book, gray-isolation, gray-virtues} to the
widespread adoption of loosely-consistent distributed key-value
stores~\cite{bigtable, pnuts, dynamo}. Fundamental to this trend was
the 2000 introduction of Brewer's CAP Theorem, which stated that a
highly available system cannot provide strong consistency guarantees
in the presence of network partitions~\cite{brewer-slides}. As
formally proven~\cite{gilbert-cap}, the CAP Theorem concerns
linearizability: the ability to read the most recent write to a data
item that is replicated across servers~\cite{herlihy-art}. Yet this
result---regarding a fairly narrow form of distributed
consistency---is often conflated with ACID database
properties~\cite{brewer-slides, hn, foundation-article}, leading to
substantial confusion about the relationships between replica
consistency, transactional isolation, and high availability. However,
transactions \textit{are} useful for programmers as evidenced by the
recent resurgence of transactional data stores such as Google's
Spanner~\cite{spanner}. Unfortunately, these systems are not highly
available in the presence of partitions~\cite{middleware-db,
  foundation-article, hstore,generalizedsnapshot, mdcc,
  krikellas-bargain, eiger, walter, calvin}.

Indeed, serializable transactions---the gold standard of traditional
ACID databases---are not achievable given a requirement for high
availability in the presence of
partitions~\cite{davidson-survey}. However, database systems have a
long tradition of providing weaker isolation and consistency
guarantees~\cite{adya, ansicritique, gray-virtues, gray-isolation,
  kemme-thesis}. Due to their concurrency and overall performance
benefits, weak isolation models are widely employed, overwhelmingly as
the default setting in today's traditional ``ACID'' and up-and-coming
``NewSQL'' databases and often as the maximum guarantee offered
(Section~\ref{sec:modernacid}). While weak isolation levels do not
provide serializability for general-purpose transactions, they are
apparently strong enough to deliver acceptable behavior to many
application programmers and are substantially stronger than the
semantics of current highly available systems. This raises a natural
question: which of these alternative semantics can be provided while
remaining highly available?

Unfortunately, the relationship of the CAP Theorem and ACID semantics
has not been well explored. We have a strong understanding of weak
isolation in the single-server context from which it
originated~\cite{adya, ansicritique, gray-isolation} and many papers
offer techniques for providing distributed
serializability~\cite{bernstein-book, spanner, daudjee-session,
  hstore, krikellas-bargain, calvin, kemme-classification} or snapshot
isolation~\cite{daudjee-snapshot,generalizedsnapshot, kemme-snapshot,
  walter}. The distributed computing, programming language, and
parallel hardware literature also contains many consistency models for
single operations on replicated objects~\cite{ ceri-mechanism,
  chen-mechanism, pnuts, herlihy-art, eiger, cac, sessionguarantees}. However, the literature has largely ignored semantic guarantees for
multiple operations operating on multiple data items in a highly
available distributed environment.

Our main contributions are as follows. We fill the gap between, on one
hand, the plethora of previously proposed isolation and consistency
models, and, on the other hand, the goal of high availability.  We
classify which, among the wide array of models, are highly available,
denoting them as {\em Highly Available Transactions (HATs)}. In doing
so, we demonstrate that many \textit{implementations} of HAT semantics
are not highly available but that their unavailability is not
\textit{inherently} required. Our investigation shows, somewhat
surprisingly, that besides serializability, only Snapshot Isolation
(SI) and Repeatable Reads (RR) are inherently unavailable, with most
other isolation levels being available. We also show that many of the
weaker data consistency models from distributed systems are
HAT-compliant.

More precisely, the reason why fairly strong models like Snapshot
Isolation (SI) and Repeatable Reads (RR) are not highly available is
that they require detecting conflicts between concurrent updates (as
needed for preventing Lost Updates or Write Skew phenomena). However,
Read Committed isolation, Transactional Atomicity, and many of the
weaker data consistency models from database and distributed systems
can be implemented in a highly available manner. For several
guarantees, such as Transactional Causal Predicate Cut Isolation, we
consider a modified form of high availability in which clients ``stick
to'' (i.e., have affinity with) at least one server (a property which
is often implicit in the distributed systems
literature~\cite{herlihy-art, eiger, cac} but which requires
\textit{explicit} consideration in a client-server replicated database
context). This sticky availability is easily implementable and widely
employed~\cite{eiger, vogels-defs} but is strictly less available than
traditional definitions of high availability.

This taxonomy allows us to determine that a wide range of semantic
guarantees can be delivered with availability in face of partitions,
but which guarantees are worthwhile in practice? We attempt to
objectively study the virtues and limitations of both classic and HAT
systems by surveying practitioner accounts and research literature,
performing experimental analyses on modern cloud infrastructure, and
analyzing representative applications for their semantic
requirements. While our results are not definitive, they suggest that
HATs offer a one to three order of magnitude latency decrease compared
to traditional distributed serializability protocols, and they can
provide acceptable semantics for a wide range of programs, especially
those with monotonic logic and commutative updates~\cite{calm, blooml,
  crdt}. HAT systems can also enforce arbitrary foreign key
constraints on multi-item updates and sometimes provide limited
uniqueness guarantees. However, particularly for programs with
non-monotonic logic, HATs may fall short, requiring
sometimes-unavailable mechanisms that are more
coordination intensive. Overall, this paper quantifies the benefits
Highly Available Transactions can offer as well as the necessary
semantic restrictions they place on applications.

>>>>

The last decade has seen a large shift in the design of large-scale
database backends, from the use of transactional
RDBMSs~\cite{bernstein-book, gray-isolation, gray-virtues} to the
widespread adoption of loosely-consistent distributed key-value
stores~\cite{bigtable, pnuts, dynamo}. A prominent event in this shift
was the 2000 introduction of Brewer's CAP Theorem, which stated that a
highly available system cannot provide strong consistency guarantees
in the presence of network partitions~\cite{brewer-slides}. As
formally proven~\cite{gilbert-cap}, the CAP Theorem concerns
linearizability: the ability to read the most recent write to a data
item that is replicated across servers~\cite{herlihy-art}. Yet the
implications of CAP---which concerns a fairly narrow form of
distributed consistency---are often conflated with the ability to
provide ACID database properties~\cite{brewer-slides, hn,
  foundation-article}; this has led to substantial confusion regarding
replica consistency, transactional isolation, and high
availability. While transactions \textit{are} useful for programmers
as evidenced by the recent resurgence of transactional data stores
such as Spanner~\cite{spanner}, these systems are unfortunately not
highly available in the presence of partitions~\cite{middleware-db,
  foundation-article, hstore,generalizedsnapshot, mdcc,
  krikellas-bargain, eiger, walter, calvin}.

Indeed, serializable transactions---the gold standard of traditional
ACID databases---are not achievable with high availability in the
presence of partitions~\cite{davidson-survey}. However, database
systems have a long tradition of providing weaker isolation and
consistency guarantees~\cite{adya, ansicritique, gray-virtues,
  gray-isolation, kemme-thesis}. Due to their concurrency and overall
performance benefits, weak isolation models are widely employed,
overwhelmingly as the default setting in today's traditional ``ACID''
and up-and-coming ``NewSQL'' databases and often as the maximum
guarantee offered (Section~\ref{sec:modernacid}). While weak isolation
levels do not provide serializability for general-purpose
transactions, they are apparently strong enough to deliver acceptable
behavior to many application programmers and are substantially
stronger than the semantics provided by current highly available
systems. This raises a natural question: which of these alternative
semantics can be provided while remaining highly available?

Unfortunately, the relationship of high availability and ACID
semantics has not been well explored. We have a strong understanding
of weak isolation in the single-server context from which it
originated~\cite{adya, ansicritique, gray-isolation} and many papers
offer techniques for providing distributed
serializability~\cite{bernstein-book, spanner, daudjee-session,
  hstore, krikellas-bargain, calvin, kemme-classification} or snapshot
isolation~\cite{daudjee-snapshot,generalizedsnapshot, kemme-snapshot,
  walter}. Additionally, the distributed computing, programming
language, and parallel hardware literature contains many
consistency models for single operations on replicated objects~\cite{
  ceri-mechanism, chen-mechanism, pnuts, herlihy-art, eiger, cac,
  sessionguarantees}. However, the literature has largely ignored
semantic guarantees for multiple operations operating on multiple data
items in a highly available distributed environment.

Our main contributions in this paper are as follows. We fill the gap
between, on one hand, the plethora of previously proposed isolation
and consistency models, and, on the other hand, the goal of high
availability.  We classify which among the wide array of models are
achievable with high availability, denoting them as {\em Highly
  Available Transactions} (HATs). In doing so, we demonstrate that many
implementations of HAT semantics are not highly available but
that their unavailability is not inherently required. Our
investigation shows, somewhat surprisingly, that besides
serializability, only Snapshot Isolation and Repeatable Reads are not
HAT-compliant, while most other isolation levels are achievable with
high availability. We also show that many weak data consistency models
from distributed systems are also HAT-compliant.

Making HAT classifications requires study of several phenomena. For
example, Snapshot Isolation (SI) and Repeatable Reads (RR) are not
HAT-compliant because they require detecting conflicts between
concurrent updates (as needed for preventing Lost Updates or Write
Skew phenomena). However, Read Committed isolation, Transactional
Atomicity, and many of the weaker data consistency models from
database and distributed systems are achievable via novel algorithms
that rely on multi-versioning and limited client-side caching. For
several guarantees, such as causal consistency with phantom prevention
and ANSI Repeatable Read, we consider a modified form of high
availability in which clients ``stick to'' (i.e., have affinity with)
at least one server (a property which is often implicit in the
distributed systems literature~\cite{herlihy-art, eiger, cac} but
which requires explicit consideration in a client-server replicated
database context). This sticky availability is easily implementable
and widely employed~\cite{eiger, vogels-defs} but is strictly less
available than traditional definitions of high availability.

This taxonomy allows us to determine that a wide range of semantic
guarantees can be delivered with availability in face of partitions,
but which guarantees are worthwhile in practice? We study the virtues
and limitations of both classic and HAT systems by surveying
practitioner accounts and research literature, performing experimental
analysis on modern cloud infrastructure, and analyzing representative
applications for their semantic requirements. While our results are
not definitive, they suggest that HATs offer a one to three order of
magnitude latency decrease compared to traditional distributed
serializability protocols, and they can provide acceptable semantics
for a wide range of programs, especially those with monotonic logic
and commutative updates~\cite{calm, blooml, crdt}. HAT systems can
also enforce arbitrary foreign key constraints on multi-item updates
and sometimes provide limited uniqueness guarantees. However,
particularly for programs with non-monotonic logic, HATs can fall
short, requiring sometimes-unavailable mechanisms that are more
coordination intensive. Overall, this paper quantifies the benefits
Highly Available Transactions can offer as well as the necessary
semantic restrictions they place on applications.

### 2013-04-23 writeup/model.tex
editing pass

Some algorithms assume a model in which the client always contacts the
same server across subsequent operations. As we will discuss in
Section~\ref{sec:hats}, clients can ensure continuity between
operations (e.g., reading their prior updates to a data item) by
maintaining affinity or ``stickiness'' with a server or set of
servers~\cite{vogels-defs}. Simultaneously maintaining availability
and stickiness entails fate-sharing between the client and its 
server: if a client's sticky server fail, or, equivalently, if the
client is partitioned from its sticky server, then it may face
unavailability. We say that a system provides \textit{sticky
  availability} if, whenever a client's operation is executed against
a correct server that has observed all of its prior operations, it eventually
receives a response, even in the presence of indefinitely long
partitions. A client may choose to become sticky available by acting
as a server itself; for example, a client might cache its reads and
writes~\cite{bolton, sessionguarantees, swift}.

Availability implies sticky availability, since availability does not
require that the contacted server has seen all the previous operations
of a client, whereas sticky availability is the special case in which
the system only guarantees correctness if the contacted server has
seen all the prior operations from the client.

@@

We have so far focused on single-object, single-operation
availability. This is standard in the distributed systems literature
(e.g., linearizable and regular register models all concern single
objects~\cite{herlihy-art}), yet the database literature largely
focuses on transactions: groups of multiple operations over multiple
objects. Accordingly, by itself, replica availability is insufficient
to describe availability guarantees for transactions. Additionally,
given the choice of \textit{commit} and \textit{abort}
responses---which signal transaction success or failure to a
client---we must carefully define transactional availability.

>>>>

Many distributed algorithms assume a model in which clients always
contact the same (sets of) server(s) across subsequent operations. As
we will discuss in Section~\ref{sec:hats}, clients can ensure
continuity between operations (e.g., reading their prior updates to a
data item) by maintaining affinity or ``stickiness'' with a server or
set of servers~\cite{vogels-defs}. Simultaneously maintaining
availability and stickiness requires fate-sharing between the client
and its server: if a client's sticky servers fail or if the client is
partitioned from its sticky servers, then it may face
unavailability. We say that a system provides \textit{sticky
  availability} if, whenever a client's operation is executed against
a correct server that has observed all of its prior operations, it
eventually receives a response, even in the presence of indefinitely
long partitions. A client may choose to become sticky available by
acting as a server itself; for example, a client might cache its reads
and writes~\cite{bolton, sessionguarantees, swift}. Any guarantee
achievable in a highly available system is achievable in a sticky high
availability system but not vice-versa.

@@

We have so far focused on single-object, single-operation
availability. This is standard in the distributed systems literature
(e.g., distributed register models such as linearizability all
concern single objects~\cite{herlihy-art}), yet the database
literature largely focuses on transactions: groups of multiple
operations over multiple objects. Accordingly, by itself, replica
availability is insufficient to describe availability guarantees for
transactions. Additionally, given the choice of \textit{commit} and
\textit{abort} responses---which signal transaction success or failure
to a client---we must take care in defining transactional availability.

### 2013-04-23 writeup/relatedwork.tex
editing pass

Weak consistency and high availability have been well
studied. Serializability has long been known to be
unachievable~\cite{davidson-survey} and Brewer's CAP Theorem has
attracted considerable attention~\cite{gilbert-cap}. The research has
begun to offer hints regarding the connection between ``weak
consistency'' and low latency~\cite{abadi-pacelc}, while several
studies examine weak isolation guarantees~\cite{adya,
  ansicritique}. However, as we have discussed, the connection between
weak consistency models in a distributed setting~\cite{calm, crdt} and
high availability is relatively unexplored. With the exception of a
preliminary workshop paper introducing HATs~\cite{hat-hotos}--- which
this work expands, providing impossibility results, a full
taxonomization, and experimental and application analysis---we believe
this paper is the first to explore this area. Nonetheless, several
studies rigorously classify non-HAT semantics: Wisemann et al. have
proposed a three parameter classification for one-copy serializable
database systems, performing a more extensive classification than our
overview in Section~\ref{sec:evaluation}, including eager database
replication techniques~\cite{kemme-classification}. Additionally
several earlier surveys discuss distributed mechanisms for achieving
one-copy serializability~\cite{wisemann-survey}, linearizability,
consistent snapshots, and recency bounds~\cite{ceri-mechanism,
  chen-mechanism}.

@@

With HATs, we seek an alternative set of semantics that are still
useful but do not violate requirements for high availability or low
latency. Recent systems proposals such as Swift~\cite{swift},
Eiger~\cite{eiger}, and Bolt-on Causal Consistency~\cite{bolton}
provide transactional causal consistency guarantees with varying
availability and represent a new class of (sticky) HAT
systems.

>>>>

Weak consistency and high availability have been well
studied. Serializability has long been known to be
unachievable~\cite{davidson-survey} and Brewer's CAP Theorem has
attracted considerable attention~\cite{gilbert-cap}. Recent work has
begun to offer hints regarding the connection between ``weak
consistency'' and low latency~\cite{abadi-pacelc}, while several
studies examine weak isolation guarantees~\cite{adya,
  ansicritique}. However, as we have discussed, the connection between
weak consistency models in a distributed setting~\cite{calm, crdt} and
high availability is relatively unexplored. With the exception of a
preliminary workshop paper discussing highly available RC and
I-CI~\cite{hat-hotos}--- which this work expands, providing additional
algorithms and impossibility results, a full taxonomization, and
experimental and application analysis---we believe this paper is the
first to explore this area. Nonetheless, several studies rigorously
classify non-HAT semantics: Wisemann et al. have proposed a three
parameter classification for one-copy serializable database systems,
performing a more extensive classification than our overview in
Section~\ref{sec:evaluation}, including eager database replication
techniques~\cite{kemme-classification}. Additionally several earlier
surveys discuss distributed mechanisms for achieving one-copy
serializability~\cite{wisemann-survey}, linearizability, consistent
snapshots, and recency bounds~\cite{ceri-mechanism, chen-mechanism}.

@@

With HATs, we seek an alternative set of semantics that are still
useful but do not violate requirements for high availability or low
latency. Recent systems proposals such as Swift~\cite{swift},
Eiger~\cite{eiger}, and Bolt-on Causal Consistency~\cite{bolton}
provide transactional causal consistency guarantees with varying
availability and represent a new class of sticky HAT systems.

\pbnote{add TA stuff}
